D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-19 ootavacuous.tex
Updates for Appendix D4
This is an extensive set of edits for Appendix D4.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This section contains several example litmus tests from the active
mind of Alan Stern, expanding on the material presented in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

@@

First, consider Listing~\ref{lst:No Invented Atomic Loads}.
At first glance, this might appear straightforward.
\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
If both dependencies are semantic dependencies, the OOTA cycle cannot
form.

@@

But is \co{P0()}'s dependency a semantic dependency?
For example, is a C++ compiler permitted to
duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
values?
If this is permitted, then the compiler would be within its rights
to transform
Listing~\ref{lst:No Invented Atomic Loads}
into
Listing~\ref{lst:Atomic Loads Invented}.

Because the values of \co{r1}, \co{r2}, and \co{r3} are all known to be
either zero or one, if the values returned by the two loads from \co{x}
differ, then one of those two values must be equal to that from \co{y}.
If the compiler is further permitted to make a runtime choice
between the two values loaded from \co{x}, it could always choose the
one that was equal to the value loaded from \co{y}, as shown by the
\co{if} statement on line~13.
This would mean that the store to \co{z} was always the constant 1,
which would permit the store to be ordered before the load from \co{y},
as shown in lines~14 and~15.
Doing so permits the OOTA-like outcome of \co{r3 == r4 == 1}.

The compiler might also drop the load from \co{y}, whose value
is now unused except by the \co{exists} clause.
However, the \co{exists} clause is considered to be observable behavior,
which means that the load from \co{y} must remain.

@@

One possible reaction to this situation is to declare that useful
C++ compilers should never duplicate nonvolatile
relaxed atomic loads.
But suppose that the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Such unused loads can easily be generated by complex macros or
	template metaprograms.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing a load, but instead merely inventing a use of
the previously discarded value in \co{r0}.

@@

This issue is not restricted to booleans.
For example, if given the code fragment shown in
Figure~\ref{lst:OOTA Cycle}
on
page~\pageref{lst:OOTA Cycle},
is the compiler permitted to act as if the code was instead as shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}?
Here, if the value loaded by line~9 is something other than 42,
the load is retried on line~11, and perhaps a third time on line~14.
If the value loaded in all three cases is something other than 42,
then line~16 stores the last value loaded, and otherwise, the constant
42 is stored by either line~18, 20, or~23.
Despite the fact that the code in both listings can be argued to be
equivalent, it is difficult to argue that a reasonable compiler
would act as if the code in
Figure~\ref{lst:OOTA Cycle}
was instead that shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.

@@

But suppose that we drop the single-thread-analysis constraint for
the moment, and suppose that the compiler is able to prove
that the only modification of a given atomic variable is to atomically
increment it, as is the case for \co{y} in
Listing~\ref{lst:Only Atomic Increment}.

@@

Is a C++ compiler within its rights to behave as if the code
shown in
Listing~\ref{lst:Only Atomic Increment}
was instead as shown in
Listing~\ref{lst:Only Atomic Increment, Optimized}?
One argument in favor is that the value of \co{y} must have
passed through the value 42 in the case where \co{r0} and \co{r1}
bracket that value.
Arguments against might cite the added overhead of the invented load
in Listing~\ref{lst:Only Atomic Increment, Optimized}, or that the
possibility that the check for the value 42 is intended for statistical
sampling.

@@

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
\item	Compilers should not invent uses for values from
	nonvolatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	nonvolatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time a compiler might be tempted to invent uses from
	values from nonvolatile relaxed atomic loads from the same
	object, the loads should instead be fused into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	Compilers that invent nonvolatile atomic loads or invent
	new uses for nonvolatile atomic loads can adversely affect
	observed behavior, for example, by introducing error into code
	attempting to do statistical sampling.
\item	What other interesting situations might come to light?
\end{enumerate}

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from x out of
the loop:

\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit i++)
    y[i] += x;
\end{verbatim}
\end{quote}

A compiler might act as if the program had instead been this:

\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit i++)
    y[i] += r1;
\end{verbatim}
\end{quote}

Note that passing \co{x} to the call to \co{do_something_with()} is important
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.

But if \co{x} is a nonvolatile atomic, then there are no worries about
data races.
In theory, the compiler could make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry.
After all, optimizations are supposed to speed things up.
This means that a compiler might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid expensive cache misses.

>>>>

This section contains several example litmus tests
expanding on the material presented in
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}.

@@

Consider Listing~\ref{lst:No Invented Atomic Loads}.
At first glance, it might appear straightforward.
\co{P0()}'s store to \co{z} clearly depends on its load from \co{y},
and \co{P1()}'s store to \co{y} clearly depends on its load from \co{z}.
If both dependencies are semantic, the OOTA cycle cannot be realized.

@@

But is \co{P0()}'s dependency truly semantic?
Imagine what might happen if a C++ compiler is permitted to
duplicate \co{P0()}'s load from \co{x} and then make use of both loads'
values.
This would allow a compiler to transform
Listing~\ref{lst:No Invented Atomic Loads}
into
Listing~\ref{lst:Atomic Loads Invented}.

Because of the comparisons against zero on lines~9 and~10 of
Listing~\ref{lst:No Invented Atomic Loads},
the values of \co{r1} and \co{r2} are known to be either zero or one.
Hence if the values of \co{r1a} and \co{r1b} on lines~9 and~10 of
Listing~\ref{lst:Atomic Loads Invented}
differ (say because \co{P2()} changes the value of \co{x}
during the time between the two loads)
then one of them must be equal to the value that would
be obtained for \co{r2}.
If the compiler is further permitted to make a choice at runtime
between the two values loaded from \co{x}, it could always choose
to use for \co{r1} the one that would be equal to \co{r2} (see line~16).
Then the value stored to \co{z} would always be one,
as shown on line~14, where the store has been moved before
the load from \co{y} on line~15 to emphasize the fact that
it does not depend on that load.
The dependency from~\co{y} to~\co{z} would not be semantic after all,
and the outcome \co{r1 == r2 == 1} would indeed be possible
(although it would not be an example of OOTA because of the lack of a
semantic dependency).

This is a simple example of a more general phenomenon.
Suppose $z$ is given by some function $f$ of $x$ and $y$.
Under what conditions can we say that the value of $z$ doesn't depend
on $y$?
The standard answer is that this happens when there are values
$z_0$ and $x_0$ such that for any $y$, we have $z_0 = f(x_0, y)$.
But if we are allowed to choose from among multiple values of $x$,
the situation gets more complicated.
Then the answer would be that there is a value $z_0$ and
a set $X$ of values for $x$ such that for any $y$, there is some
$x \in X$ with $z_0 = f(x, y)$.
In Listing~\ref{lst:Atomic Loads Invented},
$f$ is the equality function and $X$ is simply the set $\{0, 1\}$;
however, the reasoning applies in any situation where $x$ ranges
over a finite collection of possible values.
And as the example shows, the more general condition can hold in situations
where the simpler condition does not, indicating that our intuitive
notions of semantic dependency are not adequate when there can be
multiple loads of the same variable.

@@

Clearly something has gone wrong if a valid (in loose C++) transformation
like this one is capable of destroying what should be an obvious semantic
dependency.
One possible reaction is to declare that useful C++ compilers should never
invent or duplicate nonvolatile relaxed atomic loads.
But what if the initial code was as shown in
Listing~\ref{lst:Unused Extra Atomic Load}?\footnote{
	Unused loads can easily be generated by complex macros or
	template metaprograms.
	Typically compilers then remove them, but they are not obliged to.}
In this case, the transformation to
Listing~\ref{lst:Atomic Loads Invented}
would not require inventing or duplicating a load,
but instead merely inventing a use for
the previously discarded value in \co{r0}.

@@

This issue is not restricted to booleans.
For example, if given the code fragment shown in
Figure~\ref{lst:OOTA Cycle}
on
page~\pageref{lst:OOTA Cycle},
is the compiler permitted to act as if the code was instead as shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}?
Here, if the value loaded by line~9 is something other than 42,
the load is retried on line~11, and perhaps a third time on line~14.
If the value loaded in all three cases is something other than 42,
then line~16 stores the last value loaded, and otherwise, the constant
42 is stored by either line~18, 20, or~23.
Despite the fact that the code in both listings can be argued to be
equivalent, it is difficult to imagine that a reasonable compiler
would act as if the code in
Figure~\ref{lst:OOTA Cycle}
was instead that shown in
Listing~\ref{lst:Unused Inventing Non-Boolean Atomic Loads}.

@@

@@

Let us drop the single-thread-analysis constraint for
the moment, and suppose that the compiler is able to prove
that the only modification of a given atomic variable is to atomically
increment it, as is the case for \co{y} in
Listing~\ref{lst:Only Atomic Increment}.
Is it okay for a C++ compiler to transform this program into
Listing~\ref{lst:Only Atomic Increment, Optimized}?
An argument in favor is that the value of \co{y} must have
passed through the value 42 if \co{r0} and \co{r1}
bracket that value.
Arguments against might cite the added overhead of the invented load,
or the possibility that the check for the value 42 is intended for statistical
sampling.

@@

\begin{enumerate}
\item	Listing~\ref{lst:No Invented Atomic Loads}
	looks like it has an OOTA cycle, but there is in fact no
	semantic dependency.
\item	Attempts to use real-world constraints to prevent OOTA cycles
	are futile.
\item	Compilers should not invent uses for values from
	nonvolatile relaxed atomic loads.
	If it turns out that there are useful optimizations that invent
	such uses, then such optimizations must be applied only with
	careful attention to the as-if rule.
	Some might argue that inventing comparisons between a pair of
	nonvolatile relaxed atomic loads from the same object should
	be permitted only if the two loads could be reordered to be
	adjacent to each other.
\item	Any time a compiler might be tempted to invent uses for
	values of nonvolatile relaxed atomic loads from the same
	object, the loads should instead be merged into a single load.
	This reaction might be attractive to those who carefully consider
	the pointlessness of keeping two loads that return the same value
	on the one hand or the cache-miss overhead incurred when closely
	spaced loads return different values on the other.
\item	Compilers that invent nonvolatile atomic loads or invent
	new uses for nonvolatile atomic loads can adversely affect
	observed behavior, for example, by introducing error into code
	attempting to do statistical sampling.
\item	What other interesting situations might come to light?
\end{enumerate}

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.

But if \co{x} is a nonvolatile atomic, there are no worries about
data races.
In theory, if the \co{do_something_with()} call was not present
the compiler could still make the above transformation without
worry, especially given that the value loaded from \co{x} is discarded in
the case where \co{limit} is less than or equal to zero.
Except that the data-race worry is replaced by a cache-miss worry.
After all, optimizations are supposed to speed things up.
A compiler might be wise to adhere to the
same restriction that avoids data races for loads from non-atomic
variables in order to avoid potentially expensive cache misses.
